Smart Model Elimination for Efficient Automated Machine Learning

### 2024-06-16 paper.tex
draft methodology

\section{Methodology}

\subsection{Observation}
When considering a specific dataset, various aspects regarding the distribution and relationship within the data can give us clues as to which models are likely to perform best. For example, linear models tend to work better on datasets where there is a linear relationship between features. As a result, we can develop a boosting model to predict the best models based off extracted attributes from a dataset.

\begin{table}
  \caption{Attributes extracted from a dataset}
  \label{model-options-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    Attribute & Formula & Purpose\\
    \midrule
    \# of Rows & $|\mathbf{A}|_r$ & Data Size \\
    \hline
    \# of Columns & $|\mathbf{A}|_c$ & Data Size \\
    \hline
    Target Distribution &
    $\frac{\max(\text{count}_1, \text{count}_0)}{|\mathbf{A}|_r}$
    & Data Type \\
    \hline
    Numerical Columns &
    $\frac{|\mathbf{A}_{num}|}{|\mathbf{A}|_{c}}$
    & Data Type \\
    \hline
    Binary Categorical Columns &
    $\frac{|\mathbf{A}_{bin}|}{|\mathbf{A}|_{c}}$
    & Data Type \\
    \hline
    Mean \# of Distinct Values  & $\mu \left\{\text{\# of distinct values in column}\right\}$ & Data Type \\ 
    \hline
    Mean IQR  & $\mu \left\{\text{IQR of column}\right\}$ & Data Distribution \\ 
    \hline
    STD of IQR  & $\sigma \left\{\text{IQR of column}\right\}$ & Data Distribution \\
    \hline
    Mean Q1 & $\mu \left\{\text{Q1 of column}\right\}$ & Data Distribution \\
    \hline
    STD of Q1  & $\sigma \left\{\text{Q1 of column}\right\}$ & Data Distribution \\
    \hline
    Mean Q3  & $\mu \left\{\text{Q3 of column}\right\}$ & Data Distribution \\
    \hline
    STD of Q3  & $\sigma \left\{\text{Q3 of column}\right\}$ & Data Distribution \\
    \hline
    Mean Percent of Z-Score Outliers  & 
    $\mu \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & Outlier Data Points \\
    \hline
    STD of Percent of Z-Score Outliers  &
    $\sigma \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & Outlier Data Points \\
    \hline
    Mean Correlation& $\mu \left\{\text{Correlation Matrix}\right\}$ & Data linearity \\
    \hline
    STD of Correlation & $\sigma \left\{\text{Correlation Matrix}\right\}$ & Data Linearity \\
    \bottomrule
  \end{tabular}
\end{table}

\begin {figure}
\centering
\fbox{\includegraphics[width=\textwidth]{smeml-flowchart.png}}
\caption{SMEML Process}
\end{figure}
\subsection{Model Elimination}

>>>>

\section{Methodology}
When considering a specific dataset, various aspects regarding the distribution and relationship within the data can give us clues as to which models are likely to perform best. For example, linear models tend to work better on datasets where there is a linear relationship between features. As a result, we can develop a boosting model to predict the best models based off extracted attributes from a dataset. These extracted attributes are shown in Table 2.

We considered 28 different models, including all supervised and semi-supervised sklearn classification models and common boosting models such as XGBoost, Lightgbm, and Catboost.

To implement this model elimination process, we trained a boosting model on 268 kaggle datasets. We choose datasets that were medically related and focused our approach to binary classification. We extracted the attributes in Table 2 from each dataset, this is the input for the boosting model. We then ran a "dumb" framework that trained every model on each dataset. The accuracy of every model is ranked and every model is treated as a label with its value being its rank. Then we trained a multilabel regression to predict the rank of each model giving the extracted attributes. We produced a median spearman correlation of 0.7301 between the predicted rank and the actual rank and a median p-value of 1.032e-05. Furthermore, we guarantee that we can predict a top two model within the predicted top ten 90.12\% of the time.

When a user inputs a dataset into SMEML, it first extracts the attributes from the dataset. It feeds these attributes into the boosting model to get the predicted rank of each model. Since we have determined that we can predict a top two model within the predicted top ten quite confidently we can eliminate the bottom 18 models from the list. We run bayseian optimization on the top ten models to get the best hyperparamters for each model. We select the top models that have significantly better performance than other models. These models are stacked together to form the final model. This process is shown in Figure 1.
\begin{table}
  \caption{Attributes extracted from a dataset}
  \label{model-options-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    Attribute & Formula & Purpose\\
    \midrule
    \# of Rows & $|\mathbf{A}|_r$ & Data Size \\
    \hline
    \# of Columns & $|\mathbf{A}|_c$ & Data Size \\
    \hline
    Target Distribution &
    $\frac{\max(\text{count}_1, \text{count}_0)}{|\mathbf{A}|_r}$
    & Data Type \\
    \hline
    Numerical Columns &
    $\frac{|\mathbf{A}_{num}|}{|\mathbf{A}|_{c}}$
    & Data Type \\
    \hline
    Binary Categorical Columns &
    $\frac{|\mathbf{A}_{bin}|}{|\mathbf{A}|_{c}}$
    & Data Type \\
    \hline
    Mean \# of Distinct Values  & $\mu \left\{\text{\# of distinct values in column}\right\}$ & Data Type \\ 
    \hline
    Mean IQR  & $\mu \left\{\text{IQR of column}\right\}$ & Data Distribution \\ 
    \hline
    STD of IQR  & $\sigma \left\{\text{IQR of column}\right\}$ & Data Distribution \\
    \hline
    Mean Q1 & $\mu \left\{\text{Q1 of column}\right\}$ & Data Distribution \\
    \hline
    STD of Q1  & $\sigma \left\{\text{Q1 of column}\right\}$ & Data Distribution \\
    \hline
    Mean Q3  & $\mu \left\{\text{Q3 of column}\right\}$ & Data Distribution \\
    \hline
    STD of Q3  & $\sigma \left\{\text{Q3 of column}\right\}$ & Data Distribution \\
    \hline
    Mean Percent of Z-Score Outliers  & 
    $\mu \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & Outlier Data Points \\
    \hline
    STD of Percent of Z-Score Outliers  &
    $\sigma \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & Outlier Data Points \\
    \hline
    Mean Correlation& $\mu \left\{\text{Correlation Matrix}\right\}$ & Data linearity \\
    \hline
    STD of Correlation & $\sigma \left\{\text{Correlation Matrix}\right\}$ & Data Linearity \\
    \bottomrule
  \end{tabular}
\end{table}

\begin {figure}
\centering
\fbox{\includegraphics[width=\textwidth]{smeml-flowchart.png}}
\caption{SMEML Process}
\end{figure}
